---
title: "Predicting Wildlife Sighting Opportunities"
subtitle: "Hard Task — GSoC 2026 Ecotourism"
author: "Atharv Raskar"
date: today
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: show
    theme: cosmo
---

## Overview

The ecotourism package provides occurrence records for four Australian organisms
alongside daily weather data. A common question for ecotourists is: *"When and
where should I go to have the best chance of spotting this animal?"*

To answer that, we need a model that links historical sighting frequency to
weather conditions, and then projects forward to rank upcoming dates by expected
sighting count.

This document demonstrates `predict_sighting()`, a function that fits a count
regression model (Poisson or negative binomial GLM) on daily sighting totals
joined with weather covariates, and returns the top predicted days over the next
year.

```{r}
#| label: setup
#| message: false
#| warning: false

library(ecotourism)
library(dplyr)
library(ggplot2)
library(knitr)

source("predict_sighting.R")
```

## How the function works

The key steps inside `predict_sighting()`:

1. **Pick a station** — defaults to the weather station with the most sightings
   for the chosen organism.
2. **Aggregate daily counts** — count sightings per day at that station.
3. **Join with weather** — left join onto the weather table so that days with
   no sightings get an explicit zero. The model needs those zeros.
4. **Fit a GLM** — start with Poisson; if the data is overdispersed (dispersion
   ratio > 1.5), refit as negative binomial. Wildlife counts are almost always
   overdispersed due to clumped behaviour.
5. **Project forward** — build a grid of future dates with expected weather
   based on climatological averages (historical means by calendar day).
6. **Rank** — predict sighting counts on the future grid, attach confidence
   intervals, return the top N days.

Both the occurrence data and the weather data are passed as arguments — nothing
is hardcoded inside the function.

---

## Gouldian Finch

The Gouldian Finch is an endangered bird found primarily in the Top End of
Australia. During the dry season (roughly June–October), surface water
contracts and the birds gather at remaining waterholes — making them easier
to spot.

```{r}
#| label: gouldian-predict

gf_preds <- predict_sighting(gouldian_finch, weather, n_days = 5)
kable(gf_preds, digits = 2, caption = "Top 5 predicted days for Gouldian Finch")
```

```{r}
#| label: gouldian-plot
#| fig-width: 8
#| fig-height: 4.5

plot_predictions(gf_preds)
```

We'd expect the model to favour dry-season months — high temperatures, low
precipitation. That's exactly when the birds concentrate at waterholes, which
is the ecological mechanism behind the pattern.

```{r}
#| label: gouldian-temp
#| fig-width: 7
#| fig-height: 4

plot_weather_effect(gf_preds, var = "temp")
```

---

## Manta Rays

Reef Manta Rays (*Mobula alfredi*) are observed along the Queensland coastline
and at coral reef cleaning stations. Warmer water draws them to shallower
reef areas, and calmer conditions (low wind) make them more visible.

One important data issue: the manta ray dataset contains machine observations
(automated underwater cameras) which are concentrated at midnight. Filtering
to human observations gives more ecologically meaningful results.

```{r}
#| label: manta-predict

manta_human <- manta_rays |> filter(record_type == "HUMAN_OBSERVATION")

mr_preds <- predict_sighting(manta_human, weather, n_days = 5)
kable(mr_preds, digits = 2, caption = "Top 5 predicted days for Manta Rays (human obs.)")
```

```{r}
#| label: manta-plot
#| fig-width: 8
#| fig-height: 4.5

plot_predictions(mr_preds)
```

---

## Glowworms

Tasmanian Glowworms (*Arachnocampa tasmaniensis*) are bioluminescent insects
typically spotted at night in cool, humid caves and rainforest gullies. With
only 124 records in the dataset, the model has less data to work with —
which may show up as wider confidence intervals.

```{r}
#| label: glow-predict

gw_preds <- predict_sighting(glowworms, weather, n_days = 5)
kable(gw_preds, digits = 2, caption = "Top 5 predicted days for Glowworms")
```

```{r}
#| label: glow-plot
#| fig-width: 8
#| fig-height: 4.5

plot_predictions(gw_preds)
```

With only 124 sightings concentrated at a few stations, the model estimates
are less precise. The wider confidence bands reflect this honestly.

---

## Orchids

Orchids (family Orchidaceae) make up the largest dataset with over 35,000
records across Australia. Spring months (September–November) are peak
flowering season, especially in Western Australia and New South Wales.

Given the large dataset, the model has plenty of signal to work with.

```{r}
#| label: orchid-predict

or_preds <- predict_sighting(orchids, weather, n_days = 5)
kable(or_preds, digits = 2, caption = "Top 5 predicted days for Orchids")
```

```{r}
#| label: orchid-plot
#| fig-width: 8
#| fig-height: 4.5

plot_predictions(or_preds)
```

---

## Model Comparison

Let's compare the models fitted for each organism side by side.

```{r}
#| label: model-comparison

comparison <- tibble(
  Organism     = c("Gouldian Finch", "Manta Rays", "Glowworms", "Orchids"),
  Model        = c(attr(gf_preds, "model_type"),
                   attr(mr_preds, "model_type"),
                   attr(gw_preds, "model_type"),
                   attr(or_preds, "model_type")),
  Dispersion   = round(c(attr(gf_preds, "dispersion"),
                          attr(mr_preds, "dispersion"),
                          attr(gw_preds, "dispersion"),
                          attr(or_preds, "dispersion")), 2),
  AIC          = round(c(attr(gf_preds, "aic"),
                          attr(mr_preds, "aic"),
                          attr(gw_preds, "aic"),
                          attr(or_preds, "aic")), 1),
  N_obs        = c(attr(gf_preds, "n_obs"),
                   attr(mr_preds, "n_obs"),
                   attr(gw_preds, "n_obs"),
                   attr(or_preds, "n_obs")),
  Top_date     = format(c(gf_preds$date[1], mr_preds$date[1],
                           gw_preds$date[1], or_preds$date[1]), "%b %d, %Y"),
  Best_hour    = c(gf_preds$best_hour[1], mr_preds$best_hour[1],
                   gw_preds$best_hour[1], or_preds$best_hour[1])
)

kable(comparison, caption = "GLM comparison across all four organisms")
```

### What the models tell us

- **Gouldian Finch**: likely negative binomial due to clumped waterhole
  observations. Dry season months dominate. Best hour is typically early
  morning when birds visit water before the heat peaks.

- **Manta Rays**: warm-water months rank highest. Low wind speed is a
  significant predictor — calm seas = better diving visibility.

- **Glowworms**: small sample size makes this the least precise model.
  Cool, humid conditions should rank higher. Night-time observations
  only (bioluminescence is invisible in daylight).

- **Orchids**: the most data-rich model. Spring months dominate, matching
  flowering phenology. Moderate rainfall triggers growth, but not heavy
  downpours.

---

## Limitations and possible improvements

- **Future weather is approximated** using historical averages (climatological
  normals). Integrating a weather forecast API (e.g., Open-Meteo) for the
  next 7–14 days would give more accurate short-term predictions.

- **Zero inflation** may affect organisms with very sparse sightings. A
  zero-inflated negative binomial model could be explored for cases like
  glowworms.

- **Spatial component** — the current function models one station at a time.
  A multi-station model or spatial random effect could capture regional
  variation.

- **Observer effort** is a confounding factor. More sightings on weekends
  might reflect more people looking, not more animals present. The model
  does not currently account for this.

These are natural extensions that could be part of the full GSoC 2026 project
scope.
