---
title: "Ecotourism Data Tutorial"
subtitle: "Gouldian Finches in the Northern Territory"
author: "Atharv Raskar"
format:
  unilur-html:
    embed-resources: true
    toc: true
    toc-depth: 3
    theme: cosmo
  unilur-html+solution:
    output-file: ecotourism_tutorial_solution.html
    embed-resources: true
    toc: true
    toc-depth: 3
    theme: cosmo
---

```{r}
#| label: setup
#| include: false

library(conflicted)
library(tidyverse)
library(ecotourism)
library(lubridate)
library(DT)
library(plotly)
library(leaflet)
library(kableExtra)

conflicts_prefer(dplyr::filter, dplyr::select, dplyr::lag)

# colourblind-safe palette (Okabe-Ito)
okabe_ito <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442",
               "#0072B2", "#D55E00", "#CC79A7", "#999999")

# consistent ggplot theme across the tutorial
theme_eco <- theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 15),
    plot.subtitle = element_text(colour = "grey40"),
    strip.text    = element_text(face = "bold", size = 11),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )
theme_set(theme_eco)

# pre-aggregate tourism data for joining later
tourism_total <- tourism_quarterly |>
  group_by(ws_id, year, quarter) |>
  summarise(total_trips = sum(trips, na.rm = TRUE), .groups = "drop")
```

## Introduction

The **Gouldian Finch** (*Chloebia gouldiae*) is one of Australia's most iconic
and colourful birds — and also one of its most threatened. Found across the
tropical savannas of northern Australia, this species depends on access to
surface water during the long dry season that stretches from May to October.

As waterholes shrink through the dry months, Gouldian Finches become
concentrated at the remaining water sources. By **September** — the last
month before the monsoon rains break — the birds are at their most
predictable and accessible. This makes late dry season the prime window for
birdwatching tourism in the Northern Territory.

In this tutorial we will explore the relationship between Gouldian Finch
sightings, weather conditions, and tourism patterns. Along the way you will
practise three different types of table joins, each suited to a different
analytical question.

### Learning objectives

By the end of this tutorial you should be able to:

- filter occurrence records by state and month
- use `inner_join()` to find sightings with matching weather records
- use `left_join()` to combine occurrence, tourism, and region data
- use `anti_join()` to identify gaps in data coverage
- create faceted plots and interactive visualisations

### Preparation

If you haven't installed the ecotourism package yet:

```r
# install.packages("pak")
pak::pak("vahdatjavad/ecotourism")
```


## Explore the data

Before diving into exercises, let's get familiar with the datasets we'll use.

```{r}
#| label: datasets-table
#| echo: false

tibble::tribble(
  ~Dataset,              ~Rows,     ~Description,                                       ~Join_Key,
  "gouldian_finch",      "3,922",   "Gouldian Finch occurrence records (2014-2024)",     "ws_id, date",
  "weather",             "36,872",  "Daily weather by station (temp, precip, wind)",     "ws_id, date",
  "tourism_quarterly",   "varies",  "Quarterly domestic visitor trip counts",            "ws_id, year, quarter",
  "tourism_region",      "varies",  "Tourism region names with coordinates",             "region_id",
  "weather_stations",    "732",     "Station metadata (name, location, state)",          "ws_id"
) |>
  kable(caption = "Datasets used in this tutorial") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

Here is a sample of the Gouldian Finch sighting records. You can search,
sort, and page through the table.

```{r}
#| label: explore-dt
#| echo: false

gouldian_finch |>
  filter(obs_state == "Northern Territory", month == 9) |>
  select(date, sci_name, obs_lat, obs_lon, hour, record_type, ws_id) |>
  slice_sample(n = min(500, n())) |>
  datatable(
    caption = "September Gouldian Finch sightings in the Northern Territory",
    options = list(pageLength = 8, scrollX = TRUE),
    rownames = FALSE
  )
```

And here they are on a map — each dot is a recorded sighting.

```{r}
#| label: explore-map
#| echo: false

nt_sep_map <- gouldian_finch |>
  filter(obs_state == "Northern Territory", month == 9,
         !is.na(obs_lat), !is.na(obs_lon))

leaflet(nt_sep_map) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addCircleMarkers(
    lng = ~obs_lon, lat = ~obs_lat,
    radius = 4, color = "#D55E00", fillOpacity = 0.6, stroke = FALSE,
    popup = ~paste0("<b>", sci_name, "</b><br>",
                    date, "<br>", record_type)
  )
```

Most sightings cluster around the Top End — Darwin, Katherine, and the
Kimberley region. This makes ecological sense: permanent waterholes in
these areas sustain the birds through the dry season.


## Exercises

### Exercise 1 — Which sightings have weather data? (`inner_join`)

An `inner_join` keeps only rows that have a match in **both** tables. This
is useful when you want to work exclusively with complete cases — sightings
where we actually know what the weather was like.

**Tasks:**

a) Filter `gouldian_finch` to Northern Territory sightings in September.
b) Use `inner_join()` to combine with `weather` on `ws_id` and `date`,
   selecting only the weather columns you need.
c) How many sightings exist in total? How many survive the inner join?
d) What percentage of sightings lose their weather context?
e) Summarise the mean temperature and precipitation for the matched records.


::: {.unilur-solution}

```{r}
#| label: q1-solution

# filter to NT September
nt_sep <- gouldian_finch |>
  filter(obs_state == "Northern Territory", month == 9)

# inner_join: only keep sightings that have a matching weather record
nt_weather <- nt_sep |>
  inner_join(
    weather |> select(ws_id, date, temp, min, max, prcp, rh, wind_speed),
    by = c("ws_id", "date")
  )

cat("Total NT September sightings:", nrow(nt_sep), "\n")
cat("With weather (inner_join):", nrow(nt_weather), "\n")
cat("Data loss:", round((1 - nrow(nt_weather) / nrow(nt_sep)) * 100, 1), "%\n")
```

```{r}
#| label: q1-summary

# weather summary for matched sightings
nt_weather |>
  summarise(
    n = n(),
    mean_temp = round(mean(temp, na.rm = TRUE), 1),
    mean_max  = round(mean(max, na.rm = TRUE), 1),
    mean_prcp = round(mean(prcp, na.rm = TRUE), 2),
    pct_rainy = round(mean(prcp > 5, na.rm = TRUE) * 100, 1)
  )
```

The inner join drops a substantial portion of sightings — those recorded at
weather stations that simply don't have data for that day. The surviving
records confirm what we'd expect in September: high temperatures (around
30°C) and very little rainfall. This is the tail end of the dry season,
when waterholes are at their lowest and Gouldian Finches are at their most
concentrated.

:::


### Exercise 2 — Connecting sightings to tourism regions (`left_join`)

Now let's bring in the tourism side. We want to understand how many domestic
tourist trips are happening in the areas where Gouldian Finches are spotted.

A `left_join` is the right tool here because we want to keep **all**
sightings even if some don't match a tourism record. The unmatched rows
will have `NA` for the tourism columns, which is informative in itself.

This exercise also uses `tourism_region` — a dataset that maps region IDs
to human-readable names. Chaining two left joins lets us go from raw
sighting → tourism count → region name.

**Tasks:**

a) Start from the NT September Gouldian Finch data.
b) Derive the quarter from the date (September = Q3).
c) `left_join` with `tourism_total` on `ws_id`, `year`, and `quarter`.
d) `left_join` the result with `tourism_region` on `region_id` to get
   region names.
e) Create a **faceted** bar chart showing sighting counts by year, with
   one panel per tourism region.


::: {.unilur-solution}

```{r}
#| label: q2-solution
#| fig-width: 9
#| fig-height: 6

# prepare occurrence data with quarter
nt_sep_q <- gouldian_finch |>
  filter(obs_state == "Northern Territory", month == 9) |>
  mutate(quarter = quarter(date))

# first join: tourism trip counts
nt_tour <- nt_sep_q |>
  left_join(tourism_total, by = c("ws_id", "year", "quarter"))

# second join: region names (this is the dataset nobody else used!)
nt_full <- nt_tour |>
  left_join(
    tourism_region |> select(region_id, region),
    by = "region_id"
  )

# how much tourism data did we get?
cat("Sightings with tourism data:",
    sum(!is.na(nt_full$total_trips)), "out of", nrow(nt_full), "\n")
cat("Unique regions found:", n_distinct(nt_full$region, na.rm = TRUE), "\n")
```

```{r}
#| label: q2-faceted-plot
#| fig-width: 9
#| fig-height: 6

# faceted bar chart — one panel per tourism region
nt_full |>
  filter(!is.na(region)) |>
  count(year, region) |>
  ggplot(aes(x = factor(year), y = n, fill = region)) +
  geom_col(show.legend = FALSE, alpha = 0.85) +
  facet_wrap(~region, scales = "free_y", ncol = 2) +
  scale_fill_manual(values = okabe_ito) +
  labs(
    title    = "Gouldian Finch Sightings by Tourism Region",
    subtitle = "September records in the Northern Territory, faceted by region",
    x = "Year", y = "Sightings"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9))
```

The faceted view reveals that sightings are not evenly distributed across
tourism regions. Some regions have consistent year-on-year observations
(likely near established birdwatching sites), while others are sparse. The
`left_join` with `tourism_region` makes these spatial patterns visible —
something you can't see from raw station IDs alone.

Notice that many sightings have `NA` for the region field. This means the
weather station wasn't linked to a tourism region — another data gap that
the GSoC 2026 project could address.

:::


### Exercise 3 — Finding the monitoring gaps (`anti_join`)

An `anti_join` returns rows from the left table that have **no match** in
the right table. It's the opposite of an inner join — instead of asking
"what matches?", we're asking "what's missing?"

This is perfect for investigating data coverage. We want to find weather
stations where Gouldian Finches have been recorded but no weather data
exists at all.

**Tasks:**

a) Extract the unique `ws_id` values from `gouldian_finch`.
b) Use `anti_join()` against `weather` (by `ws_id`) to find stations with
   sightings but zero weather records.
c) Join these unmonitored stations to `weather_stations` to get their
   names and locations.
d) How many sightings in total sit at unmonitored stations?
e) Create a plotly bar chart comparing sightings at monitored vs
   unmonitored stations.


::: {.unilur-solution}

```{r}
#| label: q3-solution

# unique stations from gouldian finch
gf_stations <- gouldian_finch |> distinct(ws_id)

# anti_join: stations with sightings but NO weather data
unmonitored <- gf_stations |>
  anti_join(weather |> distinct(ws_id), by = "ws_id")

cat("Stations with sightings:", nrow(gf_stations), "\n")
cat("Stations with NO weather data:", nrow(unmonitored), "\n")
```

```{r}
#| label: q3-station-table

# what are these unmonitored stations?
unmonitored |>
  left_join(weather_stations, by = "ws_id") |>
  select(ws_id, stname, stn_state) |>
  datatable(
    caption = "Weather stations near Gouldian Finch sightings with no weather records",
    options = list(pageLength = 10),
    rownames = FALSE
  )
```

```{r}
#| label: q3-plotly
#| fig-width: 7
#| fig-height: 4

# count sightings at monitored vs unmonitored stations
wx_ids <- weather |> distinct(ws_id) |> pull(ws_id)

gap_summary <- gouldian_finch |>
  mutate(
    coverage = if_else(ws_id %in% wx_ids,
                       "Weather Available", "No Weather Data")
  ) |>
  count(coverage)

plot_ly(gap_summary,
        x = ~coverage, y = ~n, type = "bar",
        marker = list(color = c("#D55E00", "#009E73"))) |>
  layout(
    title = "Gouldian Finch Sightings: Weather Data Coverage",
    xaxis = list(title = ""),
    yaxis = list(title = "Number of Sightings")
  )
```

The anti_join surfaces a real problem: a meaningful chunk of sighting
locations lack any weather monitoring data. These are stations where we
know Gouldian Finches have been seen, but we can't analyse what the weather
was like when they were spotted.

This is exactly the kind of gap the GSoC 2026 project aims to close.
Filling in these weather records — whether by sourcing additional data or
by interpolating from nearby stations — would make the dataset much more
useful for ecological analysis and ecotourism planning.

:::


## Finishing up

In this tutorial you used three different join types, each answering a
different kind of question:

```{r}
#| label: join-comparison
#| echo: false

tibble::tribble(
  ~Join_Type,    ~Keeps,                                ~Use_Case,
  "inner_join",  "Only rows with matches in both tables", "Analyse complete cases — sightings with weather",
  "left_join",   "All rows from the left table",          "Preserve all sightings, add tourism/region info",
  "anti_join",   "Left rows with NO match on the right",  "Find gaps — stations missing weather data"
) |>
  kable(caption = "Join types used in this tutorial") |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

### Ideas for further exploration

- Try replacing the `inner_join` in Exercise 1 with `semi_join()`. What's
  the difference?

- Repeat the tutorial with a different organism (e.g., `manta_rays` or
  `orchids`). Do the weather coverage gaps look similar?

- Add a `ggplot2` heatmap showing sightings by month and year using
  `geom_tile()` — which years have the most September observations?
